Clustering time series based on probability distributions across temporal granularities

### 2021-10-31 main.Rmd
transformation section done

@@

Time series often have a somewhat skewed distribution. Furthermore, their ranges might vary greatly. It is helpful to do a statistical transformation on the data to bring all of them to the same range or at least normalise each series such that the CDF of the resultant variable is Gaussian. For the JS-based approaches, two data transformation techniques are utilised viz, Normal-Quantile Transform (NQT) and Robust scaling. NQT is a built-in transformation for computing $wpd$, which is the foundation of wpd-based distances.

_Robust scaling_ To get a Gaussian probability distribution for an input variable, most scaling methods remove the mean from values and divide by the standard deviation (zero mean and unit variance). Standardization may be distorted or biased if the input variable contains outliers. To address this, robust scaling methods could be utilized (value – median) / (p75 – p25)) which results in a variable with a zero mean and median, as well as a standard deviation of one, while the outliers are still there with the same relative connections to other values.

_Normal-Quantile transform_ First, the original data is ranked in ascending order and the probabilities $P(Y<=y(i)) = i/(n+1)$ are attached to $y(i)$, in terms of their ranking order. A NQT based transformation is applied by computing from a standard normal distribution a variable $\eta(i)$, which corresponds to the same probability $P(\eta< \eta(i)) = i/n+1$. By doing this, the new variables $\eta(i)$ will be marginally distributed according to standard Normal, N(0,1). NQT will make the skewed distributions bell-shaped. So it's hard to tell from the modified distribution which raw distribution was which. Also, multimodality is disguised or inverted. This, however, is not a problem for this methodology as distributions are characterised by quantiles and the order of the quantiles are reserved under NQT.

@@

- _Finding significant cyclic granularities or harmonies_

\noindent These cyclic granularities are useful for exploring repetitive patterns in time series data that get lost in the linear representation of time. It is advantageous to consider only those cyclic granularities across which there is a significant repetitive pattern for the majority of customers or noteworthy in an electricity-behavior context. In that case, when the customers are grouped, we can expect to observe some interesting patterns across the categories of the cyclic granularities considered. [@Gupta2021-hakear] proposes a way to select significant cyclic granularities and harmonies which is used for this paper.

>>>>

- _Find significant granularities or harmonies_

[@Gupta2021-hakear] proposes a method for choosing significant cyclic granularities and harmonies, which is used in this work. We define "significant" granularities as those with significant distributional differences across categories. It is better to select only those granularities because it is expected that there would be some fascinating repetitive behaviour that we are interested in studying. It is worth noting that not all of the observations in the study may have the same set of important granularities. A method for selecting a list ($S_c$) of significant granularities for all observations may be as follows:

 (a) eliminate from the comprehensive list the granularities that are inconsequential for all observations.

 (b) consider only those granularities which are significant for most observations.

In both circumstances, there will be observations for which one or a few selected granularities are uninteresting. Even in that situation, having this group of observations that show no intriguing patterns over a granularity that frequently detects patterns may be useful. In contrast, if the granularities under consideration are indeed significant for a set of observations, distinct patterns could be detected while clustering them.

@@

Time series often have a somewhat skewed distribution and their ranges might vary greatly. It
is helpful to do a statistical transformation on the data to bring all of them to the same range or normalize each series. For the JS-based approaches, two data transformation techniques are utilised viz, Normal-Quantile Transform (NQT) and Robust scaling. NQT is a built-in transformation for computing $wpd$, which is the foundation of wpd-based distances.

_Robust scaling_ The normalised $i^{th}$ observation is denoted by $v_{norm} = \frac{v_t - p_{0.50}}{p_{0.75}-p_{0.25}}$, where $v_t$ is the actual value at the $t^{th}$ time point and $p_{0.25}$, $p_{0.50}$ and $p_{0.75}$ are the $25^{th}$, $50^{th}$ and $75^{th}$ percentile of the time series for the $i^{th}$ observation. $v_{norm}$ has zero mean and median, as well as a standard deviation of one, while the outliers are still there with the same relative connections to other values.

_Normal-Quantile transform_ The raw data for all observations is individually normal-quantile transformed (NQT) [@Krzysztofowicz1997-bv], so that the transformed data follows a standard normal distribution. NQT will make the skewed distributions bell-shaped. As a result, determining which raw distribution was used is difficult using the modified distribution. Also, multimodality is disguised or inverted. This, however, is not a problem for implementation of this methodology as distributions are characterised by quantiles and the order of the quantiles is reserved under NQT.

<!-- First, the original data is ranked in ascending order and the probabilities $P(Y<=y(i)) = i/(n+1)$ are attached to $y(i)$, in terms of their ranking order. A NQT based transformation is applied by computing from a standard normal distribution a variable $\eta(i)$, which corresponds to the same probability $P(\eta< \eta(i)) = i/n+1$.By doing this, the new variables $\eta(i)$ will be marginally distributed according to standard Normal, N(0,1). -->

@@
